Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
finally finished an edit of all of section 3 (with a few holes)

A~hidden Markov model carries probabilities on its states and on its
state transitions.
Both the probabilities and the
the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.

@@

Another of us was curious to try to code Viterbi's equations
directly as recursive functions.
But like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show the memoized code, we have to confess to a small
transformation of Viterbi's equations.
The logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
Moreover, to make the file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs from the file.)
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}, which minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
Our implementation computes not only the minimum
negated log probability, called a \texttt{Score},
but also the path of states which leads to that most likely outcome. 
Because the cases for all three states are so similar, 
we~show only the code for the match state:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{viterbi}
\begin{smallverbatim}
viterbi' :: State -> Int -> Int -> (Score, [State])
viterbi' Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
                                          , trans i_m Ins
                                          , trans d_m Del 
                                          ]
                  , Mat : path )
  where (prevV, path) = viterbi prev (j-1) (i-1)
        trans t prev = prevV + tProb hmm (j-1) t
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

@@

MRFy's search works with a sequence of \emph{generations}.
A~generation is a collection of scored \emph{placements}.
We~pass the
search function a \texttt{Scorer}, which is the result of partially
applying \texttt{score} to the model.
\smallverbatiminput{scoredecl}
A~search algorithm is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallverbatiminput{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may not be any better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} considers the score of the best
placement of the trial generation, along with the history of previous
best scores.
It~also considers the \texttt{Age} of the trial generation, which is
the total number of 
generations tried.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the history of scores and the age of the current
generation. 
\end{itemize}
This interface allows for considerable variation.
For example, random-mutation hill climbing accepts only generations
with better scores than their predecessors,
while simulated annealing may accept a generation with a worse score.
A~genetic algorithm uses \texttt{nextGen} to
combine placements from the previous generation, while
random-mutation
hill climbing simply permutes a placement into a new one.
The interface \texttt{quit} makes it easy
to implement such termination criteria as a fixed number of generations
or a limit of going $k$~generations without achieving more than $E$
improvement in score.

We hope you find the code simple and straightforward:
\smallverbatiminput{search}

In MRFy, the Viterbi function is called several times (once for each 
non-beta-strand region) for each search generation.
Furthermore, in a 
population-based search such as \textit{multistart simulated annealing}, the 
scoring function (which itself calls Viterbi) is called many times per 
generation.
Beyond traditional profiling, we found parallelization to be an 
obvious way to improve runtime performance.
We found this to be trivial thanks 
to the \texttt{Control.Parallel} library.
We pass the Viterbi function to a map over all 
non-beta-strand regions of the model and query sequence, and these regions are 
independent of one another.
Similarly, we pass the scoring function to a map 
over the population of potential solutions, which are also independent.
We 
merely substituted \texttt{parmap rseq} for map.
MRFy now efficiently uses all six 
cores on an AMD workstation for a non-population-based simulated annealing 
search, and all forty-eight cores on a compute server for population-based 
searches.

Despite the beauty of some of the code we were able to write, we found several 
instances where the easiest path seemed to be to write ugly code.
The first of 
these relates to the otherwise elegant Viterbi function.
Our stochastic search 
approach calls the Viterbi function a tremendous number of times, and most of 
the program's runtime is spent in this function.
However, the Viterbi algorithm 
retains not just a score (the probability of the sequence given a state path) 
but also the state path itself.
We implemented this state path as a list of 
\texttt{Int}, and every step in the recursion conses a new element onto the 
list.
However, at each step of the stochastic search, we discard the state 
path; we are searching to optimize score only.
We only need the state path at 
the termination of the search, when we wish the program to output an alignment 
of the query sequence to the Markov random field model.
We found that 
optimizing the Viterbi function by removing this state path entirely improved 
runtime performance by nearly 50\%.
However, we still needed the Viterbi 
function to provide a state path when we called it after search finally 
terminated.
Thus, we wondered how to implement the Viterbi function so that a 
state path was optional.

In order to determine if this optimization was even worthwhile, we first wrote 
a quick and dirty approach that is shameful in any language, but easy: we 
duplicated code.
We copied \texttt{viterbi} to produce a \texttt{viterbiF} 
which did not retain state path.
Next, however, we wondered how to avoid code 
duplication.
Clearly, littering the code with conditionals would produce code 
that was ugly as well as less efficient.
We knew how we would solve this 
problem in other languages; we would have used macros in Lisp, or 
metaprogramming via \texttt{eval} in Ruby, to produce two versions of the 
function from one prototype.
We struggled, however, to solve this problem 
idiomatically in Haskell.

Rather than attempt to use \texttt{TemplateHaskell} to generate two versions of 
the Viterbi function, we wrote a higher-order function type 
\lstinline!type ScorePathCons a = a -> [a] -> [a]! and two functions conforming to that type:
\begin{smallverbatim}
consPath :: ScorePathCons a
consPath x xs = x:xs

consNoPath :: ScorePathCons a
consNoPath _ _ = []
\end{smallverbatim}

Then, depending on whether \texttt{viterbi} was called with \texttt{consPath} 
or \texttt{consNoPath} as a parameter, the path would or would not be 
allocated.
We found this higher-order function approach to be simple in code 
and in concept, and it imposes negligible run-time overhead as compared to 
having two distinct functions.

@@

Another source of ugliness resulted from adding cost-center annotations when we 
wished to profile our code.
GHC will happily add cost centers to top-level 
functions, but much of our program relies on sub-functions.
We found that 
manually adding \texttt{\{-SCC -\}} annotations to dozens of guard clauses and 
sub-functions harmed the readability of our program, such that we felt 
compelled to remove all such annotations as soon as we could.
In our prior 
experience with profiling tools such as kcachegrind, simply compiling with 
debug symbols -- rather than manually annotating many lines of code -- allowed 
the profiling tools to provide us enough information.
Without these 
annotations, both call-site and allocation profiles are relatively useless in 
GHC.
Perhaps a compile-time flag that would auto-annotate every equation -- 
even if the resulting cost center is simply named with a line number -- would 
make this task easier.

The final source of ugliness in our code came from attempts at debugging 
run-time errors.
MRFy relies on a source of input data -- the paired beta 
strands in the Markov random field -- which is represented in an awkward manner 
by the original SMURF program, and which is outside our control.
Rather than 
represent beta strands directly, SMURF represents pairings of residues, leaving 
us to infer and reconstruct beta strands.
This task was difficult due to 
unclear invariants around overlapping, doubly-paired beta strands (and it would 
have been difficult in any programming language; the difficulty was one of 
representation).
However, the debugging tools in GHC caused us some 
difficulties in determining the sources of run-time errors.
We used 
\texttt{trace} extensively, but this style of ``\texttt{printf} debugging'' 
littered our code.
Being initially unaware of the backtrace feature of the GHC 
profiler, we wrote wrappers around \texttt{Vector.slice} and \texttt{Vector.!} 
which themselves called trace, which at least kept our ``\texttt{printf} 
debugging'' less cluttered.
Nonetheless, we struggled with these runtime errors 
due to an awkward representation, and wished for debugging tools such as we 
might find in gdb, which would allow us to examine the stack when a backtrace 
occurs.

>>>>

A~hidden Markov model carries probabilities on its states and on its
state transitions.
Both the probabilities and the
the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
We~represent a state using an algebraic data type:
\begin{smallverbatim}
data HMMState = Mat | Ins | Del | Beg | End
\end{smallverbatim}

@@

Another of us was curious to try to code Viterbi's equations
directly as recursive functions.
Like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show the memoized code, we have to confess to a small
transformation of Viterbi's equations.
The logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
Moreover, to make the file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs from the file.)
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}, which minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

Our implementation computes not only the minimum
negated log probability, called a \texttt{Score},
but also the path of states which leads to that most likely outcome. 
Elsewhere, MRFy pairs not only paths but also other types of values
with scores, so we have defined a small abstraction:
\verbatiminput{vscore}
We~made \texttt{Scored} an instance of~\texttt{Ord} and of
\texttt{Functor}.

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state: 
we~choose the best solution
from among the three predecessor states; 
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

@@

MRFy's search works with a sequence of \emph{generations}.
A~generation is a collection of scored \emph{placements}.
We~pass the
search function a \texttt{Scorer}, which is the result of partially
applying \texttt{score} to the model.
\smallverbatiminput{scoredecl}
A~search algorithm is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallverbatiminput{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may not be any better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} considers the score of the best
placement of the trial generation, along with the history of previous
best scores.
It~also considers the \texttt{Age} of the trial generation, which is
the total number of 
generations tried.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the history of scores and the age of the current
generation. 
\end{itemize}
This interface provides considerable leeway for experiments.
For~example, random-mutation hill climbing accepts only generations
with better scores than their predecessors,
while simulated annealing may accept a generation with a worse score.
A~genetic algorithm uses \texttt{nextGen} to
combine placements from the previous generation, while
random-mutation
hill climbing simply permutes a placement into a new one.
The interface \texttt{quit} makes it easy
to implement such termination criteria as a fixed number of generations
or a limit of going $k$~generations without achieving more than $E$
improvement in score.

\seclabel{search}

We hope you find the search code simple and straightforward:
\smallverbatiminput{search}
\emph{[MENTION AND QUANTIFY NOAH'S SUCCESSES IN ADDING NEW HEURISTICS.]}

\subsection{Performance}

In MRFy, the \texttt{score} function calls \texttt{viterbi} several
 times
for each placement in each search generation.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel},
simply substituting
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per generation,
MRFy can keep six cores busy,
and when our search algorithm uses multiple placements per generation,
it~can use all forty-eight cores on a compute server.

In~\secref{viterbi} above, the example case for the \texttt{viterbi}
function computes not just a score but also an optimal path.
MRFy's \texttt{search} function, in \secref{search}, uses only the
score.
Even though Haskell evaluation is lazy, the \texttt{viterbi} function
is still allocating thunks.
We~tried cloning and modifying the \texttt{viterbi} code to compute
only a score, with no path, and this change improved 
run-time performance by nearly 50\%.
But when \texttt{search} finally decides on a placement, we need a
version of \texttt{viterbi} that \emph{does} compute path, so we can
extract the optimal parse for that placement.

How could we keep the performance improvement without having to
maintain two versions of \texttt{viterbi}?
In~Lisp or Ruby we would have used macros or metaprogramming,
but we were not confident of our ability to metaprogram a solution
using Template Haskell.
Instead, we replaced the primitive cons in \texttt{viterbi} with a
call to a function of the same type, which we passed~in.
Passing the primitive~``\texttt{:}'' gives us a path,
and passing \verb+\_ _ -> []+ gives us the same
speedup as the cloned and modified~code.
But~although
this trick is simple and easy to implement, 
we~have an uneasy feeling that it may work only because of undocumented
properties of GHC's inliner, which may be subject to change without
notice.

@@

In two major areas, our lack of skills or knowledge resulted in ugly
Haskell.
The first is profiling.
GHC's profiling is based on cost centers \cite{sansome-peyton-jones},
and although GHC 
will happily add cost centers to top-level 
functions, much of our code is in nested functions.
We~know of no better technique than to add cost-center annotations to
each guard clause and nested function,
but these annotations made our code so ugly that we felt compelled to
remove them as soon as possible.
And without the annotations, we found GHC's call-site and allocation
profiles did not provide enough information to be useful.
We~miss valgrind, whose callgrind profiler gives us enough information
without \emph{any} annotations in the source.

Our other ugly code came from  attempts to debug
run-time errors.
MRFy uses a Markov random field which is represented in a legacy file
format that uses a poorly documented representation---when beta
strands overlap and are doubly paired, the invariants of the
representation are not clear.
Dealing with this representation is difficult no matter what
programming language is used, and faults are to be expected.
Using GHC, we had a hard time diagnosing run-time errors.
Extensive use of 
\texttt{trace} 
littered our code,
even when moved from our main code to wrapper functions.
We~weren't aware of the backtrace feature of GHC's profiler,
and even after we discovered that feature, the results weren't good enough.
\emph{[WHY WASN'T THE BACKTRACE FEATURE GOOD ENOUGH?]}
In~August 2011, 
Lennart Augustsson said that the biggest advantage of Strict
Haskell is getting a stack trace on error,
and Simon Marlow said that he may have figured out how to track call
stacks properly in a lazy functional language.
We~can't wait.

### 2012-03-11 search.tex
finally finished an edit of all of section 3 (with a few holes)

search :: SearchStrategy -> Scorer -> [Seed]
       -> (Scored Placement, [Score])
search strat score (s:seeds) = runFrom seeds firstGen [] 0
 where
  firstGen = map score $ gen0 strat s
  runFrom :: [Seed] -> [Scored Placement] -> [Score]
          -> Age -> (Scored Placement, [Score])
  runFrom (s1:s2:seeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 score oldPop
        trialHist = (scoreOf $ minimum newPop) : oldHist
        (newPop, newHist) =
          if accept strat s2 newHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then
          (minimum newPop, newHist)
        else
          runFrom seeds newPop newHist (age + 1)

>>>>

search :: SearchStrategy -> Scorer -> [Seed]
       -> (Scored Placement, [Score])
search strat score (s0:seeds) = runFrom seeds firstGen [] 0
 where
  firstGen = map score $ gen0 strat s0
  runFrom :: [Seed] -> [Scored Placement] -> [Score]
          -> Age -> (Scored Placement, [Score])
  runFrom (s1:s2:seeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 score oldPop
        trialHist = (scoreOf $ minimum newPop) : oldHist
        (newPop, newHist) =
          if accept strat s2 newHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then
          (minimum newPop, newHist)
        else
          runFrom seeds newPop newHist (age + 1)
